---
title: "505 Final"
author: "Ran Wang,Yunfan Xu"
date: "4/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE)

library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(tidycensus)
library(viridis)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(RSocrata)
library(FNN)
library(mapview)
library(gifski)
library(RSocrata)
library(leaflet)
library(osmdata)
library(corrplot)
library(stringr)
library(car)
library(stargazer)

setwd("/Users/inordia/Desktop/UPenn搞起来/505/505_final")

root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

st_c <- st_coordinates

palette5 <- c("#d4b1c6","#c187a9","#a64d79","#83325c","#542345")

palette4 <- c("#d4b1c6","#c187a9","#a64d79","#83325c")

qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],3),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]],
                                  c(.01,.2,.4,.6,.8), na.rm=T), digits = 3))
  }
}


q5 <- function(variable) {as.factor(ntile(variable, 5))}
```

## 1 Introduction

Since the start of the COVID-19 pandemic, public transit systems in the U.S. have faced severe challenges as ridership plummeted. Recovering ridership became a key for the survival of many public transit systems. This project aims to explore factors that influence the ridership recovery of public transit systems to better design policies that foster ridership recovery, using the case of MTA subway in New York City.      

MTA experienced a shrinkage in ridership when the pandemic started, and the ridership recovery was stumbling with ups and downs throughout the time. A more disturbing trend is the rise in violence related to transit; transit crimes are up 65.5 percent over the first two weeks of 2022 compared to the same period last year. On January 15th 2022, an Asian woman was killed after being pushed in front of a moving subway train at Times Square station, and on April 12th, a man set off smoke grenades in a crowded subway car and then opened fire, injuring at least 23 people. A rise in random attacks scared off passengers in droves and was likely to worsen the stumbled ridership recovery due to the commuting shift to work-from-home. But MTA’s long-term survival depends on the return of riders and fares which make up nearly 40% of MTA’s operating revenue. Therefore, it is crucial that research looks into the impacts of various aspects of the pandemic on transit ridership recovery. This project will explore what factors influence the ridership recovery of MTA subway in post-pandemic New York.      

This project uses data from various sources including MTA weekly fare data, New York open data, OpenStreetMap, and ACS. We then developed a group of time-series regression models to explore the relationship between potential explanatory variables and ridership recovery ratios for subway stations. We identified the best model and discussed its implications and limitations.      


## 2 Data Wrangling     

Data for this project is collected from MTA, New York open data, NYCHealth, OpenStreetMap, and ACS. Specifically, we cleaned the MTA weekly fare data into the ridership recovery ratios of 2021 comparable to 2019 pre-pandemic ridership for every station complex between week 23 to week 49, approximately from June to November. We select this time frame to avoid the holiday season in December. We could not use 2022 ridership data because other data such as crime were not available for 2022. We then spatially join other data such as crime, COVID cases, and demographics to MTA stations to create the table for regression. We also calculate the distance from stations to nearest n key trip generators such as parks and hospitals, and the number of key trip generators that fall into the buffer of stations.     

```{r mta}
mta <- read.csv("mta.csv", check.names = FALSE, encoding = "UTF-8")

mta<-mta%>%
  gather(key="week",value="recovery", 3:29)%>%
  na.omit()

remote_complex <- read.csv("remoteID_complex.csv", encoding = "UTF-8")%>%
  select(remote, complex_id, line_name)

mta <- mta%>%
  left_join(remote_complex, by=c("Remote_ID"="remote"))

mta <- mta %>% 
  group_by(complex_id, week) %>% 
  mutate(recover = mean(recovery)) %>% 
  ungroup()%>%
  select(-Remote_ID, -recovery)%>% 
  distinct(complex_id, week, .keep_all = TRUE)

station <- read.csv("station.csv")%>%
  select(Complex.ID, Borough, Structure, GTFS.Latitude, GTFS.Longitude, ADA, ADA.Notes)

mta <- mta %>% 
  left_join(station, by = c("complex_id"="Complex.ID")) %>% 
  distinct(complex_id, week, .keep_all = TRUE)%>%
  st_as_sf(coords = c("GTFS.Longitude", "GTFS.Latitude"))%>%
  st_set_crs(4326) %>%
  st_transform('EPSG:2263')

mta.geo <- mta%>%
  select(complex_id, geometry) %>% 
  distinct()

```


```{r crime}

crime <- read.csv("NYPD_Complaint_Data_Current__Year_To_Date_.csv")
crime$CMPLNT_FR_DT <- mdy(crime$CMPLNT_FR_DT)

crime <-crime%>%
  mutate(year = floor_date(CMPLNT_FR_DT, unit = "year"))%>%
  filter(year== "2021-01-01")%>%
  filter(PREM_TYP_DESC == "TRANSIT - NYC SUBWAY")%>%
  mutate(week = week(CMPLNT_FR_DT),
         dotw = wday(CMPLNT_FR_DT, label=TRUE))

crime <- crime[!(is.na(crime$STATION_NAME) | crime$STATION_NAME==""), ]

crime <- crime%>%
  st_as_sf(coords = c("Longitude", "Latitude"))%>%
  st_set_crs(4326) %>%
  st_transform('EPSG:2263')

crime.agg <-
  crime%>%
  mutate(number = 1)%>%
  group_by(STATION_NAME, week)%>%
  summarise(Freq = sum(number))

crime.geo <- crime.agg%>%
  select(STATION_NAME, geometry)%>%
  distinct()

mta.buffer <- st_buffer(mta.geo, 600)
  
crime.table<- st_join(mta.buffer, crime.geo)%>%
  st_drop_geometry() %>% 
  distinct()%>%
  na.omit()

crime.agg.complex <- crime.agg%>%
  left_join(crime.table, by = "STATION_NAME")%>%
  st_drop_geometry() %>% 
  group_by(complex_id, week)%>%
  summarise(Freq = sum(Freq))%>%
  filter(week > 22)

crime.agg.complex$week <- as.character(crime.agg.complex$week)

mta <- mta%>%
  left_join(crime.agg.complex, by = c("complex_id"= "complex_id", "week"="week"))%>%
  rename(crime = Freq)%>%
  mutate(crime = replace_na(crime, 0))
  
```


```{r covid}
zip <- st_read("Modified Zip Code Tabulation Areas (MODZCTA).geojson")
covid<- read.csv("/Users/inordia/Desktop/UPenn搞起来/505/505_final/covid.csv", check.names=FALSE)
covid <- tail(covid, -6)
covid <- covid %>% 
  rename(zip = week_ending)%>%
  gather(key="date",value="covid", 2:86)

covid$date <- mdy(covid$date)

covid <- covid %>% 
  mutate(week = week(date))%>%
  left_join(zip%>%select(modzcta, pop_est, geometry), by = c("zip"="modzcta"))

covid <- covid %>% 
  st_as_sf()%>%
  st_transform('EPSG:2263')

covid.geo <- covid %>% 
  select(zip, week, covid, geometry) %>% 
  st_join(mta.geo)

covid.geo$week <- as.character(covid.geo$week)

covid.geo <- covid.geo %>% 
  st_drop_geometry() %>% 
  select(week, covid, complex_id) %>% 
  distinct()

mta <- mta %>% 
  left_join(covid.geo, by = c("complex_id"= "complex_id", "week"="week"))%>%
  distinct(complex_id, week, .keep_all = TRUE)
```


```{r amenities}
## Parks

park <- st_read("Open Space (Parks).geojson") %>% 
  st_transform('EPSG:2263')

mta <- mta %>% 
  mutate(park3_nn=nn_function(st_c(mta),st_c(st_centroid(park)),3),
         park1_nn=nn_function(st_c(mta),st_c(st_centroid(park)),1))

## OSM

q0 <- opq(bbox = c(-74.26013414099346,40.49536834449909,-73.6976408489098,40.919585276931876))

hospital <- add_osm_feature(opq = q0, key = 'amenity', value = "hospital") %>%
  osmdata_sf(.)

hospital.sf <- st_geometry(hospital$osm_points) %>%
  st_transform(4326) %>%
  st_sf() %>%
  cbind(., hospital$osm_points$name) %>%
  rename(NAME = hospital.osm_points.name)%>%
  st_transform('EPSG:2263')%>%
  dplyr::select(geometry)

mta <- mta %>% 
  mutate(hospital_nn=nn_function(st_c(mta),st_c(hospital.sf),1))

mta$hospital <-
  st_buffer(mta, 8000) %>%
  aggregate(mutate(hospital.sf, counter = 1),., sum) %>% 
  pull(counter)

mta <- mta %>% 
   mutate(hospital = replace_na(hospital, 0))


office <- add_osm_feature(opq = q0, key = 'office', value = "company") %>%
  osmdata_sf(.)

office.sf <- st_geometry(office$osm_points) %>%
  st_transform(4326) %>%
  st_sf() %>%
  cbind(., office$osm_points$name) %>%
  rename(NAME = office.osm_points.name)%>%
  st_transform('EPSG:2263')%>%
  dplyr::select(geometry)

mta$office <-
  st_buffer(mta, 8000) %>%
  aggregate(mutate(office.sf, counter = 1),., sum) %>% 
  pull(counter)

mta <- mta %>% 
  filter(recover < 1000) %>% 
  mutate(line = str_length(line_name))

```


```{r census}
demographics_ny <- get_acs(geography = "tract", variables = c(total_pop = "B01001_001",
                                                              male = "B01001_002",
                                                              female = "B01001_026",
                                                              white = "B02001_002",
                                                              black = "B02001_003",
                                                              american_indian = "B02001_004",
                                                              asian = "B02001_005",
                                                              pi = "B02001_006",
                                                              other = "B02001_007",
                                                              mixed = "B02001_008",
                                                              hispanic = "B03002_012",
                                                              means_trans_total = "B08301_001",
                                                              commute_transit = "B08301_010E",
                                                              commute_bike = "B08301_018E",
                                                              commute_walk = "B08301_019E",
                                                              commute_drivealone = "B08301_003E",
                                                              commute_carpool = "B08301_004E",
                                                              commute_WFH = "B08301_021E",
                                                              median_inc = "B19013_001",
                                                              median_inc_f = "B19113_001",
                                                              commute_total = "B08134_001",
                                                              commute_10 = "B08134_002",
                                                              commute_10_14 = "B08134_003",
                                                              commute_15_19 = "B08134_004",
                                                              commute_20_24 = "B08134_005",
                                                              commute_25_29 = "B08134_006",
                                                              commute_30_34 = "B08134_007",
                                                              commute_35_44 = "B08134_008",
                                                              commute_45_59 = "B08134_009",
                                                              commute_60 = "B08134_010",
                                                              occupation_total = "B08124_001",
                                                              mbas = "B08124_002", #management, business, art, and science
                                                              service = "B08124_003",
                                                              sale_office = "B08124_004",
                                                              ncm = "B08124_005", #natural resources, construction, maintain
                                                              ptmm = "B08124_006", #production, transportation
                                                              military = "B08124_007"
), 
year=2019, state=36, county=061, geometry=T, output="wide", survey = "acs5")


demographics_ny$pct_m = demographics_ny$maleE / demographics_ny$total_popE
demographics_ny$pct_f = demographics_ny$femaleE / demographics_ny$total_popE
demographics_ny$pct_w = demographics_ny$whiteE / demographics_ny$total_popE
demographics_ny$pct_b = demographics_ny$blackE / demographics_ny$total_popE
demographics_ny$pct_ai = demographics_ny$american_indianE / demographics_ny$total_popE
demographics_ny$pct_a = demographics_ny$asianE / demographics_ny$total_popE
demographics_ny$pct_pi = demographics_ny$piE / demographics_ny$total_popE
demographics_ny$pct_other = demographics_ny$otherE / demographics_ny$total_popE
demographics_ny$pct_mix = demographics_ny$mixedE / demographics_ny$total_popE
demographics_ny$pct_his = demographics_ny$hispanicE / demographics_ny$total_popE
demographics_ny$pct_transit = demographics_ny$commute_transit / demographics_ny$means_trans_totalE
demographics_ny$pct_commute_20 = (demographics_ny$commute_10E + demographics_ny$commute_10_14E + demographics_ny$commute_15_19E ) / demographics_ny$commute_totalE
demographics_ny$pct_commute_45 = (demographics_ny$commute_20_24E + demographics_ny$commute_25_29E + demographics_ny$commute_30_34E + demographics_ny$commute_35_44E) / demographics_ny$commute_totalE
demographics_ny$pct_commute_45plus = (demographics_ny$commute_45_59E + demographics_ny$commute_60E) / demographics_ny$commute_totalE
demographics_ny$pct_mbas = demographics_ny$mbasE / demographics_ny$occupation_totalE
demographics_ny$pct_service = demographics_ny$serviceE / demographics_ny$occupation_totalE
demographics_ny$pct_sale_office = demographics_ny$sale_officeE / demographics_ny$occupation_totalE
demographics_ny$pct_ncm = demographics_ny$ncmE / demographics_ny$occupation_totalE
demographics_ny$pct_ptmm = demographics_ny$ptmmE / demographics_ny$occupation_totalE
demographics_ny$pct_military = demographics_ny$militaryE / demographics_ny$occupation_totalE

demographics_ki <- get_acs(geography = "tract", variables = c(total_pop = "B01001_001",
                                                              male = "B01001_002",
                                                              female = "B01001_026",
                                                              white = "B02001_002",
                                                              black = "B02001_003",
                                                              american_indian = "B02001_004",
                                                              asian = "B02001_005",
                                                              pi = "B02001_006",
                                                              other = "B02001_007",
                                                              mixed = "B02001_008",
                                                              hispanic = "B03002_012",
                                                              means_trans_total = "B08301_001",
                                                              commute_transit = "B08301_010E",
                                                              commute_bike = "B08301_018E",
                                                              commute_walk = "B08301_019E",
                                                              commute_drivealone = "B08301_003E",
                                                              commute_carpool = "B08301_004E",
                                                              commute_WFH = "B08301_021E",
                                                              median_inc = "B19013_001",
                                                              median_inc_f = "B19113_001",
                                                              commute_total = "B08134_001",
                                                              commute_10 = "B08134_002",
                                                              commute_10_14 = "B08134_003",
                                                              commute_15_19 = "B08134_004",
                                                              commute_20_24 = "B08134_005",
                                                              commute_25_29 = "B08134_006",
                                                              commute_30_34 = "B08134_007",
                                                              commute_35_44 = "B08134_008",
                                                              commute_45_59 = "B08134_009",
                                                              commute_60 = "B08134_010",
                                                              occupation_total = "B08124_001",
                                                              mbas = "B08124_002", #management, business, art, and science
                                                              service = "B08124_003",
                                                              sale_office = "B08124_004",
                                                              ncm = "B08124_005", #natural resources, construction, maintain
                                                              ptmm = "B08124_006", #production, transportation
                                                              military = "B08124_007"), 
                           year=2019, state=36, county=047, geometry=T, output="wide", survey = "acs5")

demographics_ki$pct_m = demographics_ki$maleE / demographics_ki$total_popE
demographics_ki$pct_f = demographics_ki$femaleE / demographics_ki$total_popE
demographics_ki$pct_w = demographics_ki$whiteE / demographics_ki$total_popE
demographics_ki$pct_b = demographics_ki$blackE / demographics_ki$total_popE
demographics_ki$pct_ai = demographics_ki$american_indianE / demographics_ki$total_popE
demographics_ki$pct_a = demographics_ki$asianE / demographics_ki$total_popE
demographics_ki$pct_pi = demographics_ki$piE / demographics_ki$total_popE
demographics_ki$pct_other = demographics_ki$otherE / demographics_ki$total_popE
demographics_ki$pct_mix = demographics_ki$mixedE / demographics_ki$total_popE
demographics_ki$pct_his = demographics_ki$hispanicE / demographics_ki$total_popE
demographics_ki$pct_transit = demographics_ki$commute_transit / demographics_ki$means_trans_totalE
demographics_ki$pct_commute_20 = (demographics_ki$commute_10E + demographics_ki$commute_10_14E + demographics_ki$commute_15_19E ) / demographics_ki$commute_totalE
demographics_ki$pct_commute_45 = (demographics_ki$commute_20_24E + demographics_ki$commute_25_29E + demographics_ki$commute_30_34E + demographics_ki$commute_35_44E) / demographics_ki$commute_totalE
demographics_ki$pct_commute_45plus = (demographics_ki$commute_45_59E + demographics_ki$commute_60E) / demographics_ki$commute_totalE
demographics_ki$pct_mbas = demographics_ki$mbasE / demographics_ki$occupation_totalE
demographics_ki$pct_service = demographics_ki$serviceE / demographics_ki$occupation_totalE
demographics_ki$pct_sale_office = demographics_ki$sale_officeE / demographics_ki$occupation_totalE
demographics_ki$pct_ncm = demographics_ki$ncmE / demographics_ki$occupation_totalE
demographics_ki$pct_ptmm = demographics_ki$ptmmE / demographics_ki$occupation_totalE
demographics_ki$pct_military = demographics_ki$militaryE / demographics_ki$occupation_totalE


demographics_qu <- get_acs(geography = "tract", variables = c(total_pop = "B01001_001",
                                                              male = "B01001_002",
                                                              female = "B01001_026",
                                                              white = "B02001_002",
                                                              black = "B02001_003",
                                                              american_indian = "B02001_004",
                                                              asian = "B02001_005",
                                                              pi = "B02001_006",
                                                              other = "B02001_007",
                                                              mixed = "B02001_008",
                                                              hispanic = "B03002_012",
                                                              means_trans_total = "B08301_001",
                                                              commute_transit = "B08301_010E",
                                                              commute_bike = "B08301_018E",
                                                              commute_walk = "B08301_019E",
                                                              commute_drivealone = "B08301_003E",
                                                              commute_carpool = "B08301_004E",
                                                              commute_WFH = "B08301_021E",
                                                              median_inc = "B19013_001",
                                                              median_inc_f = "B19113_001",
                                                              commute_total = "B08134_001",
                                                              commute_10 = "B08134_002",
                                                              commute_10_14 = "B08134_003",
                                                              commute_15_19 = "B08134_004",
                                                              commute_20_24 = "B08134_005",
                                                              commute_25_29 = "B08134_006",
                                                              commute_30_34 = "B08134_007",
                                                              commute_35_44 = "B08134_008",
                                                              commute_45_59 = "B08134_009",
                                                              commute_60 = "B08134_010",
                                                              occupation_total = "B08124_001",
                                                              mbas = "B08124_002", #management, business, art, and science
                                                              service = "B08124_003",
                                                              sale_office = "B08124_004",
                                                              ncm = "B08124_005", #natural resources, construction, maintain
                                                              ptmm = "B08124_006", #production, transportation
                                                              military = "B08124_007"), 
                           year=2019, state=36, county=081, geometry=T, output="wide", survey = "acs5")

demographics_qu$pct_m = demographics_qu$maleE / demographics_qu$total_popE
demographics_qu$pct_f = demographics_qu$femaleE / demographics_qu$total_popE
demographics_qu$pct_w = demographics_qu$whiteE / demographics_qu$total_popE
demographics_qu$pct_b = demographics_qu$blackE / demographics_qu$total_popE
demographics_qu$pct_ai = demographics_qu$american_indianE / demographics_qu$total_popE
demographics_qu$pct_a = demographics_qu$asianE / demographics_qu$total_popE
demographics_qu$pct_pi = demographics_qu$piE / demographics_qu$total_popE
demographics_qu$pct_other = demographics_qu$otherE / demographics_qu$total_popE
demographics_qu$pct_mix = demographics_qu$mixedE / demographics_qu$total_popE
demographics_qu$pct_his = demographics_qu$hispanicE / demographics_qu$total_popE
demographics_qu$pct_transit = demographics_qu$commute_transit / demographics_qu$means_trans_totalE
demographics_qu$pct_commute_20 = (demographics_qu$commute_10E + demographics_qu$commute_10_14E + demographics_qu$commute_15_19E ) / demographics_qu$commute_totalE
demographics_qu$pct_commute_45 = (demographics_qu$commute_20_24E + demographics_qu$commute_25_29E + demographics_qu$commute_30_34E + demographics_qu$commute_35_44E) / demographics_qu$commute_totalE
demographics_qu$pct_commute_45plus = (demographics_qu$commute_45_59E + demographics_qu$commute_60E) / demographics_qu$commute_totalE
demographics_qu$pct_mbas = demographics_qu$mbasE / demographics_qu$occupation_totalE
demographics_qu$pct_service = demographics_qu$serviceE / demographics_qu$occupation_totalE
demographics_qu$pct_sale_office = demographics_qu$sale_officeE / demographics_qu$occupation_totalE
demographics_qu$pct_ncm = demographics_qu$ncmE / demographics_qu$occupation_totalE
demographics_qu$pct_ptmm = demographics_qu$ptmmE / demographics_qu$occupation_totalE
demographics_qu$pct_military = demographics_qu$militaryE / demographics_qu$occupation_totalE


demographics_br <- get_acs(geography = "tract", variables = c(total_pop = "B01001_001",
                                                              male = "B01001_002",
                                                              female = "B01001_026",
                                                              white = "B02001_002",
                                                              black = "B02001_003",
                                                              american_indian = "B02001_004",
                                                              asian = "B02001_005",
                                                              pi = "B02001_006",
                                                              other = "B02001_007",
                                                              mixed = "B02001_008",
                                                              hispanic = "B03002_012",
                                                              means_trans_total = "B08301_001",
                                                              commute_transit = "B08301_010E",
                                                              commute_bike = "B08301_018E",
                                                              commute_walk = "B08301_019E",
                                                              commute_drivealone = "B08301_003E",
                                                              commute_carpool = "B08301_004E",
                                                              commute_WFH = "B08301_021E",
                                                              median_inc = "B19013_001",
                                                              median_inc_f = "B19113_001",
                                                              commute_total = "B08134_001",
                                                              commute_10 = "B08134_002",
                                                              commute_10_14 = "B08134_003",
                                                              commute_15_19 = "B08134_004",
                                                              commute_20_24 = "B08134_005",
                                                              commute_25_29 = "B08134_006",
                                                              commute_30_34 = "B08134_007",
                                                              commute_35_44 = "B08134_008",
                                                              commute_45_59 = "B08134_009",
                                                              commute_60 = "B08134_010",
                                                              occupation_total = "B08124_001",
                                                              mbas = "B08124_002", #management, business, art, and science
                                                              service = "B08124_003",
                                                              sale_office = "B08124_004",
                                                              ncm = "B08124_005", #natural resources, construction, maintain
                                                              ptmm = "B08124_006", #production, transportation
                                                              military = "B08124_007"), 
                           year=2019, state=36, county=005, geometry=T, output="wide", survey = "acs5")

demographics_br$pct_m = demographics_br$maleE / demographics_br$total_popE
demographics_br$pct_f = demographics_br$femaleE / demographics_br$total_popE
demographics_br$pct_w = demographics_br$whiteE / demographics_br$total_popE
demographics_br$pct_b = demographics_br$blackE / demographics_br$total_popE
demographics_br$pct_ai = demographics_br$american_indianE / demographics_br$total_popE
demographics_br$pct_a = demographics_br$asianE / demographics_br$total_popE
demographics_br$pct_pi = demographics_br$piE / demographics_br$total_popE
demographics_br$pct_other = demographics_br$otherE / demographics_br$total_popE
demographics_br$pct_mix = demographics_br$mixedE / demographics_br$total_popE
demographics_br$pct_his = demographics_br$hispanicE / demographics_br$total_popE
demographics_br$pct_transit = demographics_br$commute_transit / demographics_br$means_trans_totalE
demographics_br$pct_commute_20 = (demographics_br$commute_10E + demographics_br$commute_10_14E + demographics_br$commute_15_19E ) / demographics_br$commute_totalE
demographics_br$pct_commute_45 = (demographics_br$commute_20_24E + demographics_br$commute_25_29E + demographics_br$commute_30_34E + demographics_br$commute_35_44E) / demographics_br$commute_totalE
demographics_br$pct_commute_45plus = (demographics_br$commute_45_59E + demographics_br$commute_60E) / demographics_br$commute_totalE
demographics_br$pct_mbas = demographics_br$mbasE / demographics_br$occupation_totalE
demographics_br$pct_service = demographics_br$serviceE / demographics_br$occupation_totalE
demographics_br$pct_sale_office = demographics_br$sale_officeE / demographics_br$occupation_totalE
demographics_br$pct_ncm = demographics_br$ncmE / demographics_br$occupation_totalE
demographics_br$pct_ptmm = demographics_br$ptmmE / demographics_br$occupation_totalE
demographics_br$pct_military = demographics_br$militaryE / demographics_br$occupation_totalE


demographics_ri <- get_acs(geography = "tract", variables = c(total_pop = "B01001_001",
                                                              male = "B01001_002",
                                                              female = "B01001_026",
                                                              white = "B02001_002",
                                                              black = "B02001_003",
                                                              american_indian = "B02001_004",
                                                              asian = "B02001_005",
                                                              pi = "B02001_006",
                                                              other = "B02001_007",
                                                              mixed = "B02001_008",
                                                              hispanic = "B03002_012",
                                                              means_trans_total = "B08301_001",
                                                              commute_transit = "B08301_010E",
                                                              commute_bike = "B08301_018E",
                                                              commute_walk = "B08301_019E",
                                                              commute_drivealone = "B08301_003E",
                                                              commute_carpool = "B08301_004E",
                                                              commute_WFH = "B08301_021E",
                                                              median_inc = "B19013_001",
                                                              median_inc_f = "B19113_001",
                                                              commute_total = "B08134_001",
                                                              commute_10 = "B08134_002",
                                                              commute_10_14 = "B08134_003",
                                                              commute_15_19 = "B08134_004",
                                                              commute_20_24 = "B08134_005",
                                                              commute_25_29 = "B08134_006",
                                                              commute_30_34 = "B08134_007",
                                                              commute_35_44 = "B08134_008",
                                                              commute_45_59 = "B08134_009",
                                                              commute_60 = "B08134_010",
                                                              occupation_total = "B08124_001",
                                                              mbas = "B08124_002", #management, business, art, and science
                                                              service = "B08124_003",
                                                              sale_office = "B08124_004",
                                                              ncm = "B08124_005", #natural resources, construction, maintain
                                                              ptmm = "B08124_006", #production, transportation
                                                              military = "B08124_007"), 
                           year=2019, state=36, county=085, geometry=T, output="wide", survey = "acs5")

demographics_ri$pct_m = demographics_ri$maleE / demographics_ri$total_popE
demographics_ri$pct_f = demographics_ri$femaleE / demographics_ri$total_popE
demographics_ri$pct_w = demographics_ri$whiteE / demographics_ri$total_popE
demographics_ri$pct_b = demographics_ri$blackE / demographics_ri$total_popE
demographics_ri$pct_ai = demographics_ri$american_indianE / demographics_ri$total_popE
demographics_ri$pct_a = demographics_ri$asianE / demographics_ri$total_popE
demographics_ri$pct_pi = demographics_ri$piE / demographics_ri$total_popE
demographics_ri$pct_other = demographics_ri$otherE / demographics_ri$total_popE
demographics_ri$pct_mix = demographics_ri$mixedE / demographics_ri$total_popE
demographics_ri$pct_his = demographics_ri$hispanicE / demographics_ri$total_popE
demographics_ri$pct_transit = demographics_ri$commute_transit / demographics_ri$means_trans_totalE
demographics_ri$pct_commute_20 = (demographics_ri$commute_10E + demographics_ri$commute_10_14E + demographics_ri$commute_15_19E ) / demographics_ri$commute_totalE
demographics_ri$pct_commute_45 = (demographics_ri$commute_20_24E + demographics_ri$commute_25_29E + demographics_ri$commute_30_34E + demographics_ri$commute_35_44E) / demographics_ri$commute_totalE
demographics_ri$pct_commute_45plus = (demographics_ri$commute_45_59E + demographics_ri$commute_60E) / demographics_ri$commute_totalE
demographics_ri$pct_mbas = demographics_ri$mbasE / demographics_ri$occupation_totalE
demographics_ri$pct_service = demographics_ri$serviceE / demographics_ri$occupation_totalE
demographics_ri$pct_sale_office = demographics_ri$sale_officeE / demographics_ri$occupation_totalE
demographics_ri$pct_ncm = demographics_ri$ncmE / demographics_ri$occupation_totalE
demographics_ri$pct_ptmm = demographics_ri$ptmmE / demographics_ri$occupation_totalE
demographics_ri$pct_military = demographics_ri$militaryE / demographics_ri$occupation_totalE

ny = bind_rows(demographics_ny, demographics_ki, demographics_qu, demographics_br, demographics_ri)

ny.use <- ny %>% 
  select(-ends_with("M")) %>% 
  select(total_popE, pct_transit, median_incE, median_inc_fE,
         pct_mbas, pct_service, pct_sale_office, pct_military, ptmmE, ncmE,
         pct_w, occupation_totalE, geometry) %>% 
  mutate(pct_ptmm = ptmmE / occupation_totalE,
         pct_ncm = ncmE / occupation_totalE,
         pct_work = occupation_totalE/ total_popE) %>% 
  select(-ptmmE, -ncmE, -occupation_totalE) %>% 
  st_transform('EPSG:2263')


mta <- mta %>% 
  st_join(ny.use)
```

## 3 Exploratory Analysis

### 3.1 Dependent Variable    

The dependent variable for this project is the ridership recovery rate at station level by week. The density plot below shows that the distribution of the dependent variable is a normal distribution with some outliers on the right.

```{r dependent}
plot(density(mta$recover, na.rm=TRUE))

summary(mta$recover)
```

### 3.2 Spatial Autocorrelation

The dependent variable has a spatial autocorrelation process in which stations with higher or lower recovery ratios tend to cluster together. As the map below shows, stations in Brooklyn and Bronx have higher average recovery rates compared to stations in lower Manhattan.

```{r average}
mta <- mta %>% 
  group_by(complex_id) %>% 
  mutate(ave_recover = mean(recover)) %>% 
  ungroup()

summary(mta$ave_recover)

ggplot()+
  geom_sf(data = zip,
          fill = "antiquewhite1", color = "grey75")+
  geom_sf(data = mta %>% filter(week < 29), 
          shape = 21,
          aes(
            size = ave_recover,
            fill = ave_recover
          ),alpha = 1, color = "transparent", show.legend = "point") +
  scale_size_continuous(
    range = c(1,6),
    breaks = c(0.2423, 0.3882, 0.4444, 0.4852, 0.9608),
    labels = qBr(mta, "ave_recover"),
    name = "Recovery Ratio 2021/2019")+
  scale_fill_stepsn(
    colors = palette5,
    breaks = c(0.2423, 0.3882, 0.4444, 0.4852, 0.9608),
    guide = FALSE)+
  labs(title = "Graduated Symbol Map of Average Recovery Rate", subtitle = "MTA subway stations 2019/2021")+
  mapTheme()+
  guides(size = guide_legend(override.aes = list(fill = palette5)))
```

### 3.3 Time Fluctuation

The dependent variable also shows fluctuations throughout time. The plot below shows the system-wide average recovery ratio across the study period and there’s fluctuations–at the highest point of 0.6 and the lowest point of 0.35– that this project attempts to understand.


```{r average time}
ggplot()+
  geom_sf(data = zip,
          fill = "antiquewhite1", color = "grey75")+
  geom_sf(data = mta %>% filter(week < 29), 
          shape = 21,
          aes(
            size = recover,
            fill = recover
          ),alpha = 1, color = "transparent", show.legend = "point") +
  scale_size_continuous(
    range = c(1,6),
    breaks = c(0.01823, 0.33193, 0.40567, 0.46728, 6),
    labels = qBr(mta %>% filter(week < 29), "recover"),
    name = "Recovery Ratio 2021/2019")+
  scale_fill_stepsn(
    colors = palette5,
    breaks = c(0.01823, 0.33193, 0.40567, 0.46728, 6),
    guide = FALSE)+
  labs(title = "Graduated Symbol Map of Recovery Rate", subtitle = "MTA subway stations 2019/2021")+
  facet_wrap(~week)+
  mapTheme() +
  guides(size = guide_legend(override.aes = list(fill = palette4)))

 mta %>% 
  group_by(week)%>%
  summarise(rec = mean(recover)) %>% 
  st_drop_geometry() %>% 
  ggplot(aes(x=week, y=rec, group=1))+
  geom_line()+
  ylim(0.3, 0.5)+
   labs(title="System-wide Average Ridership Recovery Rate",
        subtitle = "MTA subway stations 2019/2021",
        x="Week", y="Recovery Rate")+
  plotTheme()
```

### 3.4 Lag Variables

Considering that some factors such as COVID can have a lagging effect on the subway ridership, we then develop the time-lag feature based on the recovery ratio, crimes, and COVID cases of each station in the previous time period. For example, a time lag of one week for COVID is the COVID cases of that station’s ZIP area a week ago. The correlation between lag variables and the dependent variable is shown in the table below. The high correlation between ridership recovery rate lag variables and the dependent variable suggests that better ridership recovery for a station in previous weeks corresponds to higher ridership recovery rate in later time periods. The correlation decreases as time lag is longer, suggesting that the lagging effects fade as the time goes.    

```{r lag}
 mta <- 
   mta %>% 
   arrange(complex_id, week) %>% 
   mutate(lagrecover= dplyr::lag(recover,1),
          lag2recover= dplyr::lag(recover,2),
          lag3recover = dplyr::lag(recover,3),
          lag4recover= dplyr::lag(recover,4),
          lagcrime = dplyr::lag(crime,1),
          lag2crime = dplyr::lag(crime,2),
          lagcovid = dplyr::lag(covid,1),
          lag2covid = dplyr::lag(covid,2),
          lag3covid = dplyr::lag(covid,3),
          lag4covid = dplyr::lag(covid,4))

 plotData.lag <-
   as.data.frame(mta)%>%
   dplyr::select(starts_with("lag"), recover) %>%
   gather(Variable, Value, -recover) %>%
   mutate(Variable = fct_relevel(Variable, "lagrecover", "lag2recover",
                                 "lag3recover", "lag4recover", "lagcrime", "lag2crime",
                                 "lagcovid", "lag2covid", "lag3covid", "lag4covid"))
 correlation.lag <-
   group_by(plotData.lag, Variable) %>%
   summarize(correlation = round(cor(Value, recover, use = "complete.obs"), 2)) %>%
   kable(caption = "MTA Stations Recovery Ratio") %>%
   kable_styling("striped", full_width = F)
 
 correlation.lag
```


### 3.5 Feature Engineering

```{r feature engineer}
 plot(density(mta$crime, na.rm=TRUE))
 plot(density(log(mta$crime)))
 
 mta <- mta %>% 
   mutate(crimelog = log(crime+1))

 plot(density(mta$covid, na.rm=TRUE))
 plot(density(log(mta$covid %>% na.omit())))
 
 mta <- mta %>% 
   mutate(logcovid = log(covid+1))
 
 mta[is.na(mta)] <- 0
```


### 3.6 Correlation

We log-transformed some variables to make their distribution more normal and more suitable for linear regression. As the plots below show, log-transformed crime and COVID variables have more normal distribution.   

```{r cor}
 mta.cor <- mta %>% 
   st_drop_geometry() %>% 
   select(-Station, -week, -complex_id, -line_name, -Borough, -Structure, -ADA, -ADA.Notes, 
          -ave_recover) %>% 
   na.omit() %>% 
   cor()

 corrplot(mta.cor, type="upper", order="hclust",
          tl.col="black", tl.srt=45, diag=FALSE)
```

## 4 Modeling

### 4.1 Variable Selection

For selecting the final variables, we construct a basic time-series regression model using all the collected variables against the MTA ridership recovery to determine which of them are statistically significant. The basic model indicates that some variables such as hospital location, have a p-value less than 0.1, and some variables like the park location have a p-value less than 0.05, while variables like percentage of production, transportation, and management workers have p-values less than 0.01. These p-values which are less than 0.1 prove that the corresponding variables are statistically significant to our research. Moreover, we conduct a backward selection to better select the variables. The result of backward selection suggests similar significant variables as the selection based on p-value, which includes the percentage of workers of certain types of occupations, station structure, median family income, lagged Covid data, and lagged recovery data.

```{r base reg}
reg.base <- lm(recover ~ Borough+Structure+ADA+covid+park3+park1+
             hospital_nn+hospital+office+crimelog+lagrecover+
             lag2recover+lag3recover+lag4recover+lagcrime+
             lag2crime+lagcovid+lag2covid+lag3covid+lag4covid+line+
               total_popE+pct_transit+median_incE+median_inc_fE+
               pct_mbas+pct_service+pct_sale_office+pct_military+
               pct_w+pct_ptmm+pct_ncm+pct_work+logcovid, data=mta)
 summary(reg.base)
 
```

```{r backward}
 step(lm(recover ~ Borough+Structure+ADA+covid+park3+park1+
           hospital_nn+hospital+office+crimelog+lagrecover+
           lag2recover+lag3recover+lag4recover+lagcrime+
           lag2crime+lagcovid+lag2covid+lag3covid+lag4covid+line+
           total_popE+pct_transit+median_incE+median_inc_fE+
           pct_mbas+pct_service+pct_sale_office+pct_military+
           pct_w+pct_ptmm+pct_ncm+pct_work+logcovid, data=mta),
      direction="backward")
```

### 4.2 Final Model

Combined the result of the p-value and backward selection, we have the following variables selected for constructing statistical models：

* **Structure**: the structure of the station       
* **Park3**: the average distance from a station to its nearest three parks        
* **Hospital_nn**: the distance from a station to its nearest hospital        
* **Hospital**: how many hospitals are within a 8000ft buffer around the station        
* **Lagrecover**: recovery rate one week ago      
* **Lag2recover**: recovery rate two weeks ago      
* **Lag3recover**: recovery rate three weeks ago       
* **Lag4recover**: recovery rate four weeks ago       
* **Lagcovid**: COVID cases per 10,000 people in the station’s ZIP Code Area one week ago       
* **Lag2covid**: COVID cases per 10,000 people in the station’s ZIP Code Area two weeks ago
* **Lag4covid**: COVID cases per 10,000 people in the station’s ZIP Code Area four weeks ago           
* **Line**: the number of subway lines running through the station       
* **Pct_transit**: percentage of workers commute by transit in the station’s census tract        
* **median_inc_fE**: median family income in the station’s census tract       
* **Pct_sale_office**: percentage of people with sale and office occupations in the station’s census tract        
* **Pct_ptmm**: percentage of people with production and transportation occupations in the station’s census tract       
* **Pct_wor*k**: percentage of working population in the station’s census tract       
* **Logcovid**: log of COVID cases per 10,000 people in the station’s ZIP Code Area       

The first time-series regression model contains only the different structural features of stations and key trip generators like hospitals and parks. The results indicate station features of elevated and opencut are statistically significant to the MTA ridership recovery with p-values less than 0.01, and that the ridership will increase by 4% and 6.6% respectively. A viaduct is also a statistically significant feature for stations with a p-value less than 0.05, that each viaduct station will reflect a 3.4% in the ridership recovery. The number of hospitals and subway lines, and the average distance to the nearest 3 parks are also significant in this model that each increase in these variables will result in a 0.04% decrease, 1.3% decrease, and 0.001% increase in the recovery respectively. The model has an R-square value of 0.043.

The second model adds the lagged covid and recovery data to the previous model. It has an R-square of 0.217, which shows an improvement from the first one. In this model, the station features of elevated and viaduct are no longer statistically significant with p-values larger than 0.1, but the number of hospitals and subway lines, the average distance to the nearest 3 parks, and opencut stations remain significant. This model indicates that all the lag data are statistically significant with -p-values less than 0.01. For example, for the recovery data, each increase in the recovery one week ago will reflect a 25% increase in the ridership recovery of the current week, and each increase in the recovery two weeks ago will reflect a 17.6% increase in recovery in the current week. For the Covid data, each increase in Covid cases per 10,000 people of the zip code area of a station one week ago will result in a 0.02% increase in the ridership recovery, and the increase two weeks ago will result in a 0.01% decrease in the recovery.

We add demographic data to the second model to construct the next model, which includes the percentage of commuters using public transit, median family income, and the percentages of the working population and different types of occupation. All the significant variables in the second model remain vital in this model, and some of the new variables express their importance in the result. With a p-value less than 0.01, each increase in the percent of commuters using public transit will result in a 3.9% increase in the ridership recovery. Although each increase in the percent of the working population will result in a 1% decrease in the ridership recovery, each increase in the percentage of workers in the production, transportation, and management industries will result in a 12.3% recovery of the MTA ridership, and each increase in the percentage of workers in sales and office jobs will reflect a 4% increase in the recovery. The model has an R-square value of 0.23 which means it explains 23% of the data and becomes the highest R-square value of all the models. Also, the ANOVA test of the three models indicates this one to be the best model. Meanwhile, the ANOVA test of this model shows that each variable has improved the model, and the result of the VIF test determines low correlations between the variables. Thus, we choose this model as our final model for explaining the ridership recovery of MTA.


```{r reg, results='asis'}
#reg1 only static data of station features and key trip generators
reg1 <- lm(recover ~ Structure + park3 + hospital_nn + hospital+ line , data=mta %>% na.omit())

#reg2 static data of station features and key trip generators + lag data
reg2 <- lm(recover ~ Structure + park3 + hospital_nn + hospital + lagrecover + 
              lag2recover + lag3recover + lag4recover + lagcovid + lag2covid + 
              lag4covid + line+ logcovid, data=mta %>% na.omit())

#reg3 static data of station features, key trip generators, and demographics + lag data

reg3 <- lm(recover ~ Structure + park3 + hospital_nn + hospital + lagrecover + 
               lag2recover + lag3recover + lag4recover + lagcovid + lag2covid + 
               lag4covid + line + pct_transit + median_inc_fE + pct_sale_office + 
               pct_ptmm  + pct_work + logcovid, data=mta %>% na.omit())

 stargazer(reg.base, reg1, reg2, reg3, type ="html", font.size = "small", single.row = TRUE)
 
 anova(reg1, reg2, reg3)
 
```

```{r reg6 test}
 anova(reg3)
 vif(reg3)
 plot(reg3)
```

## 5 Summary

According to the final model, an opencut station will have more riders compared with stations with other features, and more numbers of hospitals will reflect a slight decrease in the recovery. The increase of Covid cases will decrease ridership three weeks later is possible due to the hysteresis of policy changes that the corresponding mask requirement and passenger flow limitations will be implemented weeks after the Covid cases have increased. The most significant finding is the relationship between the recovery and type of occupations. The recovery will decrease with each increase in the percentage of the working population, mainly because of the work-from-home policy after the pandemic, however, the more percentages of occupations that require workers to be on-site, such as production, transportation, and sales, will reflect a significant increase to the MTA ridership recovery. In conclusion, our research determines that the post-pandemic ridership recovery of MTA is significantly influenced by the station structure, location of key trip generators, lag data of Covid cases and recoveries, and the types of occupations.
